---
title: 'EES 3310/5310 Lab #3'
subtitle: "Exercises with the MODTRAN Model"
author: "put your name here"
date: 'Lab: Mon. Sept. 10. Due: Mon. Sept. 17.'
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE)

# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
data_dir = "data"
script_dir = "scripts"

library(pacman)
p_load(zoo, xml2, tidyverse, stringr)

theme_set(theme_bw(base_size = 15))

source('scripts/utils.R', chdir = T)
source('scripts/format_md.R', chdir = T)
source('scripts/modtran.R', chdir = T)

```
# Chapter 4 Exercises

## Exercise 4.1: Methane

Methane has a current concentration of 1.7 ppm in the atmosphere and
is doubling at a faster rate than CO~2~.

a) Would an additional 10 ppm of methane in the atmosphere have a larger or smaller
  impact on the outgoing IR flux than an additional 10 ppm of CO~2~ at current
  concentrations?

```{r ex_4_1_a, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

Suggestion:

* Run MODTRAN in the default configuration (400 ppm CO~2~ and 1.7 ppm methane)
* Run MODTRAN with an extra 10 ppm of CO~2~ and the normal amount of methane.
* Run MODTRAN with the normal amount of CO~2~ and an extra 10 ppm of methane.

What would you look at from the three runs to figure out whether 10 ppm of methane
or 10 ppm of CO~2~ had the greater effect?


b) Where in the spectrum does methane absorb? What concentration does it take to
  begin to saturate the absorption in this band? Explain what you are looking 
  at to judge when the gas is saturated.
  
    **Note:** See the suggestions in the instructions for this lab.

```{r ex_4_1_b, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

  
c) Would a doubling of methane have as great an impact on the heat balance as a 
   doubling of CO~2~?

```{r ex_4_1_c, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.


d) What is the "equivalent CO~2~" of doubling atmospheric methane? That is to say,
   how many ppm of CO~2~ would lead to the same change in outgoing IR radiation
   energy flux as doubling methane? What is the ratio of ppm CO~2~ change to 
   ppm methane change?

```{r ex_4_1_d, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.


## Exercise 4.2: CO~2~ (Graduate students only)

a) Is the direct effect of increasing CO~2~ on the energy output at the top of
   the atmosphere larger in high latitudes or in the tropics?

    For each atmosphere, first record $I_{\text{out}}$ with CO~2~ at 400 ppm
    and then record the change when you increase CO~2~ to 800 ppm.

```{r ex_4_2_a, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.


b) Set pCO~2~ to an absurdly high value of 10,000 ppm. You will see a spike
   in the CO~2~ absorption band. What temperature is this light coming from? 
   Where in the atmosphere do you think this comes from?

    Now turn on clouds and run the model again. Explain what you see.
    Why are night-time temperatures warmer when there are clouds?

```{r ex_4_2_b, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.


## Exercise 4.3: Water vapor

Our theory of climate presumes that an increase in the temperature at ground
level will lead to an increase in the outgoing IR energy flux at the top of the
atmosphere.


a) How much extra outgoing IR would you get by raising the temperature of the 
   ground by 5&deg;C? What effect does the ground temperature have on the 
   shape of the outgoing IR spectrum and why?

```{r ex_4_3_a, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

b) More water can evaporate into warm air than into cool air. Change the
   model settings to hold the water vapor at constant relative humidity 
   rather than constant vapor pressure (the default), calculate the change
   in outgoing IR energy flux for a 5&deg;C temperature increase.
   Is it higher or lower? Does water vapor make the Earth more sensitive to
   CO~2~ increases or less sensitive?

```{r ex_4_3_b, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

c) Now see this effect in another way. 

    * Starting from the default base case, record the total outgoing 
      IR flux. 

    * Now double pCO2. The temperature in the model stays the
      same (that's how the model is written), but the outgoing IR flux
      goes down.

    * Using constant water vapor pressure, adjust the temperature offset
      until you get the original IR flux back again. Record the change in
      temperature
    
    * Now repeat the exercise, but holding the relative humidity fixed
      instead of the water vapor pressure.
    
    * The ratio of the warming when you hold relative humidity fixed
      to the warming when you hold water vapor pressure fixed is the 
      feedback factor for water vapor. What is it?

```{r ex_4_3_c, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

# Chapter 5 Exercise

## Exercise 5.2: Skin Height

a) Run the MODTRAN model in using the "Tropical" atmosphere, without clouds, and with
   present-day pCO~2~ (400 ppm). Use the ground temperature reported by the model to calculate
   $\varepsilon \sigma T_{\text{ground}}^4$, the heat flux emitted by the ground.
   Assume $\varepsilon = 1$, and I have already provided the value of the 
   Stefan-Boltzmann constant $\sigma$, as the R variable `sigma_sb`,
   which equals `r format_md(sigma_sb, digits = 3)`.
   (I defined it in the script "utils.R", which I loaded in the "setup" chunk
   in the RMarkdown document).

    Next, look at the outgoing heat flux at the top of the atmosphere (70 km) 
    reported by the MODTRAN model. Is it greater or less than the heat flux
    that you calculated was emitted by the ground?

```{r ex_5_2_a, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

b) Use the outgoing heat flux at the top of the atmosphere to calcuate the
   skin temperature (use the equation 
   $I_{\text{out}} = \varepsilon \sigma T_{\text{skin}}^4)$).
   What is the skin temperature, and how does it compare to the ground 
   temperature and the temperature at the tropopause, as reported by the 
   MODTRAN model?
   
    Assuming an environmental lapse rate of 6K/km, and using the 
    skin temperature that you calculated above, and the ground temperature
    from the model, what altitude would you expect the skin height to be?

```{r ex_5_2_b, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

c) Double the CO~2~ concentration and run MODTRAN again. Do not adjust the
   ground temperature. Repeat the calculations from (b) of the skin
   temperature and the estimated skin height.

    What is the new skin temperature? What is the new skin height?

```{r ex_5_2_c, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

d) Put the CO~2~ back to today's value, but add cirrus clouds, using the
   "standard cirrus" value for the clouds. Repeat the calculations from (b) of
   the skin temperature and the skin height.
   
    What is the new skin temperature? What is the new skin height?
    Did the clouds or the doubled CO~2~ have a greater effect on the
    skin height?

```{r ex_5_2_d, include=TRUE, message=FALSE}
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.


